Can historical data predict population responses to climate change experiments?

### 2016-12-16 precip/manuscript/precip_ms.tex
adding more manuscript edits to intro and methods

\begin{enumerate}
\item Climate is thought to be an important driver of population ecology however it can be difficult to determine which aspects of climate affect populations and at which stages in an organisms life-cycle.  Moreover, there have been few tests of whether statistical models fit to observational data on climate and populations can accurately predict the future responses of populations to natural or experimentally imposed climate conditions.   
\item Using observational data collected from the 1920's to the 1950's from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species. We tested whether statistical models based on these observational data could predict how these species respond to a experimentally imposed drought and irrigation from 2006 to 2016. 
\\item We established sixteen new plots in the same location as the historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers. The original plots were also monitored and served as ambient climate controls. In order to describe the demographic effects of the climate manipulations, we used all data, observational and experimental, to fit a set of statistical models describing each species survival, growth and recruitment in response to local competitive interactions and the effects of the climate manipulation experiment. Next we fit another set of models to only the observational data that used temperature and soil moisture as covariates to describe annual variation in the vital rates.  Finally, we used the second set of models fitted only to the observational data, to predict the effects of the experimental treatments on each species vital rates and used the vital rates to parameterize an individual based population model and predict one step ahead cover in each of the experimental plots.   
\item The experimental drought and irrigation treatments successfully lowered and increased soil moisture respectively.  Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots. At the level of individual vital rates, experimental drought reduced \textit{Hesperostipa comata} and \textit{Poa secunda} growth and survival. Drought increased and irrigation decreased \textit{Artemisia tripartita} growth. Recruitment rates for all three grasses were lower in both the drought and irrigation plots than in the control plots. 
\item Vital rate models fit to the observational data correctly predicted significant increases in X and X and significant decreaes in X and X.  However, the historical models predicted significant changes in X X and X that were not observed in the experiment.  Moreover, the experiment led to a significant change in X that was not predicted by the historical data. 
\item Our individual based population model more accurately predicted changes in the population growth rates of \textit{} in the experimental plots than a model without climate effects.  However for the other X species predictions made from the climate model were less accurate than models without climate.   
\item \emph{Synthesis}: Out of all four species examined, climate covariates only appeared to increase our predictive accuracy for one species: \textit{Hesperostipa comata}. In general the relatively weak responses to the experimental treatments agrees with the low explanatory power of climate covariates in explaining the vital rates and population dynamics of these species in this system. The highly variable environment and water limited conditions may select for species with a conservative non-responsive strategy to variation in climate both natural and experimentally imposed.

@@

Climate is widely considered one of the most powerful external forces driving changes in species abundance across space and time \citep{teller_linking_2016}.  The effects of climate on populations and ecosystems are most apparent at the largest scales in space and time: climate determines the distribution of ecosystems, treelines and the the range limits of many species \citep{}, while The paleoecological record shows that long-term climate change lead to changes in species abundance and changes in species' ranges \citep{davis}. Nevertheless, for many species it is difficult to determine how annual climate variation affects populations and demographic rates \citep{ehrlen_advancing_2016}. Observational data on species performance across years with varying climate can provide some information on how climate might affect population abundance \citep{dalgleish_climate_2010,ehrlen_advancing_2016}. However, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}. Understanding the effects of climate on populations is increasingly important, however, if we are to understand and predict the effects of climate change on populations. If models fit to historical data can be used to accurately predict the effects of experimental climate manipulations, it would be strong confirmation that the climate effects they describe are real and will hold in the future \citep{}.

@@

Linking annual climate variation to demographic performance in plants is a high priority for building population models for plants that can predict their future response to climate change \citep{crone_ability_2013,ehrlen_advancing_2016}. However, despite the clear signs that climate drives net primary productivity at the ecosystem level as and variation in individual growth rate in trees and other woody species, as well as the germination and fecundity of annual plants, there are few studies that have linked the effects of interannual climate variation to population models for plants, and fewer still that have shown that these population models can be used to accurately predict future plant responses to short term climate variation. Adler et al. \citep{adler byrn and leiker} showed that population models based on historically observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland. Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.  However, another three species, showed responses to the experimental conditions that were not well predicted by historical observations.  The authors suggested that limited replication in the historical data for two of these species and changing competitive conditions in the community may have led to the poor predictions.

The effects of annual climate variation on species in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least X different studies since

Section on climate models in this system.

To test whether our models, which are based on observational data, accurately predict the effects of natural climate variation, we conducted a precipitation manipulation experiment in a sagebrush steppe. We studied the responses of three perennial grass species and a shrub species to natural climate variation over 10 years and experimental drought and irrigation treatments over four years. We addressed two research questions: First, how much do the  survival, growth or recruitment of our target species differ between the precipitation manipulation treatments? If our experiment does affect species vital rates we interpret that as strong evidence that variation in precipitation should have an effect on populations. Our second question is whether statistical models parameterized from historical observational data can accurately predict the effects of modern climate variation and precipitation manipulation on species performance? If models based on observational effectively capture the effects of climate on species performance, they should also predict the effects of natural climate variation in the contemporary dataset as well as the effects of precipitation treatments in the experiment.

@@

The U.S. Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level. During the period of data collection (1926 – 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July). The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}. These four species, the focus of our models, accounted for over 70\% of basal cover (grasses) and 60\% of canopy cover (shrubs and forbs).

Scientists at the USSES established 26 1-m$^2$ quadrats between 1926 and 1932. Eighteen quadrats were distributed among four ungrazed exclosures, and eight were distributed in two paddocks grazed at medium intensity spring through fall. All quadrats were located on similar topography and soils. In most years until 1957, all individual plants in each quadrat were mapped using a pantograph (Blaisdell 1958). The historical data set is public and available online \citep{zachmann_mapped_2010}. In 2007, we located 14 of the original quadrats, all of which are inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method.

We extracted data on survival, growth, and recruitment from the mapped quadrats based on plants' spatial locations. Our approach tracks genets, which may be composed of multiple polygons, as they fragment and/or coalesce. Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years \citep{lauenroth_demography_2008}. We modeled vital rates using data from 22 year-to-year transitions between 1926 and 1957, and nine year-to-year transitions from 2007 to 2016. Only four quadrats were observed for the first two transitions in the 1920's, while at least 14 quadrats were observed for all subsequent transitions.

\subsection*{Precipitation experiment}
In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we selected locations for an additional 16 quadrats for the precipitation experiment. We avoided plots falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}. New plots were established in pairs, and one plot per pair was randomly assigned to either the precipitation reduction or the precipitation addition treatment. We mapped the quadrats in June, 2011 and then built the rainfall shelters and set-up the irrigation systems in the fall of 2011. We used a rain-out shelter and automatic irrigation design described in \citep{gherardi, yahdjian_rainout_2002}. Each rain-out shelter covered an area of 2.5 by 2 m and consisted of transparent acrylic shingles held up over the plot to channel 50\% of incoming rainfall off of the plot and into 75 l reservoirs. The collected water was automatically pumped out of the reservoirs and sprayed onto paired irrigation treatment plots. Pumping was initiated automatically with float switches that were triggered when water levels in the reservoirs were approximately 20 l, or equivalently irrigation was triggered once for every 6 mm of rainfall collected. We disconnected the irrigation pumps in late fall each year and re-connected them in April.  The drought shelters remained in place throughout the year.

We monitored soil moisture and air temperature in four of the precipitation experiment plot pairs using Decagon Devices (Pulman Washington) 5TM and EC-5 soil moisture sensors and 5TE temperature sensors.  We installed two soil moisture sensors in each monitored plot at 5 cm and two at 25 cm deep in the soil.  Air temperature was measured underneath the roofing of the shelter at 30 cm above ground. We also installed sensors in a nearby area receiving ambient rainfall to serve as a measure of un-manipulated control conditions. Data were logged automatically every four hours.

We conducted a simple statistical analysis to determine the effect of the rain-out shelters and irrigation on the observed soil moisture. We averaged soil moisture readings by day plot group  and then fit a simple linear model to the soil moisture data to determine the effects

We conducted a simple statistical analysis of differences in cover trends on treatment plots. The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$. We tested for the effect of precipitation treatment on this measure of cover change with a linear mixed effects model in the \textit{lme4} package \citep{Bates2015} of \textit{R}, including year and quadrat as random effects. We considered the precipitation treatment effects to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.

@@

We expected that our precipitation manipulation experiment would have effects on plants through the mechanism of altering available soil moisture during the growing season.  Because we do not have direct measurements of soil moisture for each year of observed plant cover, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton 1978}. We used an enhanced version of soilwat that has recently been developed for use in semi-arid shrubland ecosystems \citep{Bradford}. SOILWAT uses daily weather data, ecosystem specific vegetation properties and site specific soil properties to estimate water balance processes. SOILWAT specifically estimates rainfall interception by vegetation, evaporation of intercepted water, snow melt and snow redistribution, infiltration into the soil, percolation through the soil, bare-soil evaporation, transpiration from each soil layer, and drainage. We parameterized SOILWAT with the generic sagebrush steppe vegetation parameters and site specific soil texture and bulk density data. We used daily weather data collected at the USSES from 1925 until the present as weather forcing data for the SOILWAT predictions. The USSES climate station (station id: GHCND:USC00102707) is located within 2 km of the research plots.  We downloaded daily and monthly tmin, tmax, and precipitation data from the National Climate Data Centers online database.

Because the

We averaged daily soil moisture predictions for 5, 25 and 30 cm soil layers, and daily mean temperature, and then aggregated these quarterly.

>>>>

\begin{enumerate}
\item Climate is an important driver of population ecology however there have been few tests of whether observational data that links population performance with climate variation can be used to predict the responses of populations to experimentally imposed climate conditions.   
\item Using longterm historical observational data from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species. We tested whether statistical models based on these observational data could be used to predict how these species respond to a five year drought and irrigation experiment conducted from 2011 to 2016. 
\\item We established sixteen new plots at the same field cite as the original historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers. The original plots were also monitored and served as ambient climate controls. In order to describe the demographic effects of the climate manipulations, we used all data, observational and experimental, to fit a set of statistical models describing each species survival, growth and recruitment in response to local competitive interactions and the effects of the climate manipulation experiment. Next we fit another set of models to only the observational data that used seasonal soil moisture covariates to describe annual variation in the vital rates.  Finally, we used the observation-based models to predict the effects of the experimental treatments on each species vital rates. We also used an individual based population model to predict one year ahead changes in population size in the experimental plots.   
\item The experimental drought and irrigation treatments successfully lowered and increased soil moisture respectively.  Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots. At the level of individual vital rates, experimental drought reduced growth of \textit{Hesperostipa comata} and growth and survival of \textit{Poa secunda}. Drought increased, while irrigation decreased, growth of the shrub \textit{Artemisia tripartita}. 
\item Vital rate models fit to the observational data correctly predicted significant increases in X and X and significant decreaes in X and X.  However, the historical models predicted significant changes in X X and X that were not observed in the experiment.  Moreover, the experiment led to a significant change in X that was not predicted by the historical data. 
\item Our individual based population model more accurately predicted changes in the population growth rates of \textit{} in the experimental plots than a model without climate effects.  However for the other X species predictions made from the climate model were less accurate than models without climate.   
\item \emph{Synthesis}: Out of all four species examined, climate covariates only appeared to increase our predictive accuracy for one species: \textit{Hesperostipa comata}. In general the relatively weak responses to the experimental treatments agrees with the low explanatory power of climate covariates in explaining the vital rates and population dynamics of these species in this system. The highly variable environment and water limited conditions may select for species with a conservative non-responsive strategy to variation in climate both natural and experimentally imposed.

@@

Climate is widely considered one of the most powerful external forces driving changes in species abundance across space and time \citep{teller_linking_2016}.  The effects of climate on populations and ecosystems are most apparent at the largest scales in space and time: climate determines the distribution of ecosystems, treelines and the the range limits of many species \citep{}, while the recent historical and the paleoecological record shows that long-term climate change leads to changes in species range limits \citep{parmesan, davis}. Understanding and predicting the effects of climate on populations is an increasingly important goal if we are to understand and predict the effects of climate change on earth's ecosystems.

Unfortunately, for many species it is difficult to determine how annual climate variation affects populations and demographic rates \citep{teller_linking_2016, ehrlen_advancing_2016}. Observational data on species performance across years with varying climate can provide some information on how climate might affect population abundance \citep{dalgleish_climate_2010,ehrlen_advancing_2016}. However, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}. Generating predictions from observational historical data for the future novel conditions created by climate change is fraught with risk. Climate change will not only change mean annual temperature and precipitation, but also affect annual variation in these measures, and possibly the covariance between them \citep{}. Models based on the historical response of populations to annual climate variation, will therefore be extrapolating beyond the range of observed climate variation when they are used to generate predictions in the future \citep{}. If models fit to historical data can be used to accurately predict the effects of experimental climate manipulations, especially manipulations that generate extreme variation in climate, it would be strong confirmation that the climate effects they describe are real and will hold in the future \cite{adler_can_the_past_2014}.

@@

Linking annual climate variation to demographic performance in plants is a high priority for building population models for plants that can predict their future response to climate change \citep{crone_ability_2013,ehrlen_advancing_2016}. However, despite the clear signs that climate drives net primary productivity at the ecosystem level and the variation in individual growth rate in trees and other woody species, as well as the germination and fecundity of annual plants, there are few studies that have connected the effects of interannual climate variation to population models for plants, and fewer still that have tested whether these population models can be used to accurately predict the future responses of plants to short term climate variation. Adler et al. \citep{adler byrn and leiker} showed that population models based on historically observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland. Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.  However, another three species, showed responses to the experimental conditions that were not well predicted by historical observations.  The authors suggested that limited replication in the historical data for two of these species and changing competitive conditions in the community may have led to the poor predictions.

The demography and competitive interactions between three dominant grass species and a dominant shrub species at in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least X different studies since 2005. X of these studies report significant effects of historical climate variation on the vital rates and overall population growth of these species. Both precipitation and temperature have been shown to have strong species-specific effects on this system.  Although past studies used different statistical models and methods for variable selection they indicate X.  This well-studied system offers the ideal opportunity to test whether the associations between annual climate variation and plant demography we observe in long-term observational data can be used to predict the future responses of plant populations to climate change.

In this study, we report how the four dominant plant species at the USSES responded to a five year drought and irrigation experiment and use the results to address two research questions: first, how much do the growth, recruitment and survival of our target species differ between the precipitation manipulation treatments? If our experiment does affect species vital rates we interpret that as strong evidence that variation in precipitation should have an effect on populations. Second, we test whether statistical models parameterized from observational data only can accurately predict each species response to the experimental precipitation manipulation? If models based on observational effectively capture the effects of climate on species performance, they should also predict the effects of precipitation treatments in the experiment.

@@

The U.S. Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level. During the period of data collection (1926 – 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July). The vegetation is dominated by a shrub, \textit{Artemisia tripartita}, and three perennial C3 grasses: \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}. These dominant species account for over 70\% of basal cover and 60\% of canopy cover at this site.

Scientists at the USSES established 26 1-m$^2$ quadrats between 1926 and 1932. Eighteen quadrats were distributed among four ungrazed exclosures, and eight were distributed in two paddocks grazed at medium intensity spring through fall. All quadrats were located on similar topography and soils. In most years until 1957, all individual plants in each quadrat were mapped using a pantograph (Blaisdell 1958). The historical data set is public and available online \citep{zachmann_mapped_2010}. In 2007, we located 14 of the original quadrats, all of which are inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method. Daily temperature and precipitation has been monitored throughout this period at a climate station located at the USSES headquarters (station id: GHCND:USC00102707) which located within 2 km of the research plots.  We downloaded daily and monthly tmin, tmax, and precipitation data from the National Climate Data Centers online database.

We extracted data on survival, growth, and recruitment from the mapped quadrats based on plants' spatial locations. Our approach tracks genets representing individual plants.  For the shrub, each genet is associated with the basal position of a stem.  For the bunchgrasses, each genet represents a spatially distinct polygon in the mapped quadrat. These genets may fragment and/or coalesce over time. Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years \citep{lauenroth_demography_2008}. We modeled vital rates using data from 21 year-to-year transitions between 1929 and 1957, and four year-to-year transitions from 2007 to 2011.

\subsection*{Precipitation experiment}
In spring 2011, we selected locations for an additional 16 quadrats for the precipitation experiment. We located these in a large exclosure containing six of the historical permanent quadrats.  We avoided plots falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}. New plots were established in pairs, and one plot per pair was randomly assigned to either the precipitation reduction or the precipitation addition treatment. We mapped the quadrats in June, 2011 and then built the rainfall shelters and set-up the irrigation systems in the fall of 2011. We used a rain-out shelter and automatic irrigation design described in \citep{gherardi, yahdjian_rainout_2002}. Each rain-out shelter covered an area of 2.5 by 2 m and consisted of transparent acrylic shingles held up over the plot to channel 50\% of incoming rainfall off of the plot and into 75 l reservoirs. The collected water was pumped out of reservoirs and sprayed onto paired irrigation treatment plots. Pumping was initiated automatically with float switches that were triggered when water levels in the reservoirs were approximately 20 l, or equivalently irrigation was triggered once for every 6 mm of rainfall collected. We disconnected the irrigation pumps in late fall each year and re-connected them in April.  The drought shelters remained in place throughout the year.

We monitored soil moisture and air temperature in four of the precipitation experiment plot pairs using Decagon Devices (Pulman Washington) 5TM and EC-5 soil moisture sensors and 5TE temperature sensors.  We installed two soil moisture sensors in each monitored plot at 5 cm and two at 25 cm deep in the soil.  Air temperature was measured underneath the roofing of the shelter at 30 cm above ground. For each pair of manipulated plots, we also installed sensors in a nearby area to measure ambient rainfall and temperature. Data were logged automatically every four hours. We augmented automatic monitoring of the climate in these plots with by taking direct measurements of soil moisture with a handheld EC-5 soil moisture sensor around each plot at X different times in the spring and summer between 2012 and 2016.

Since we only collected automatic soil moisture and temperature data in half of the 16 experimental plots, we conducted a simple statistical analysis to determine the average effect of the rain-out shelters and irrigation treatments on the observed observed plots and then applied these average effects to all the plots in the experiment. We expected that differences in the soil moisture between drought irrigation plots might become more apparent after and during periods of rainfall we first classified the daily climate during the experimental period into dry and rainy days. Rainy days were defined as any days when average temperatures were above 3 degrees C and precipitation was over x mm. The days after a day classified as rainy were also classified as rainy. We then averaged datalogger readings by day and plot within each plot group (one drought, one irrigation and one ambient plot each). For each plot group, we then subtracted the drought and irrigation readings from the ambient measurements. We then modeled the average deviations in soil moisture and temperature in the experimental treatments as a function of season, raininess and their interaction. We used the parameters from this model to describe the average effects of the experimental treatments on soil moisture in all 16 experimental plots in the study

We conducted a simple statistical analysis of differences in cover trends on treatment plots. The response variable for this analysis was change in cover at the quadrat scale measured at the start of the experiment in spring 2011 prior to manipulation, to the last year of the experiment in 2016. We calculated this change in cover as $log(Cover_{t+1}/Cover_t)$ per plot. We tested for the effect of precipitation treatment on this change in cover with a linear model in \textit{R}. We considered the precipitation treatment effects to be significant when??.

@@

We expected that our precipitation manipulation experiment would have effects on plants through the mechanism of altering available soil moisture during the growing season.  Because we do not have direct measurements of soil moisture for each year of observed plant cover in the historical record, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton 1978}. We used an enhanced version of soilwat that has recently been developed for use in semi-arid shrubland ecosystems \citep{Bradford}. SOILWAT uses daily weather data, ecosystem specific vegetation properties and site specific soil properties to estimate water balance processes. SOILWAT specifically estimates rainfall interception by vegetation, evaporation of intercepted water, snow melt and snow redistribution, infiltration into the soil, percolation through the soil, bare-soil evaporation, transpiration from each soil layer, and drainage. We parameterized SOILWAT with the generic sagebrush steppe vegetation parameters and site specific soil texture and bulk density data. We used daily weather data collected at the USSES from 1925 until the present as weather forcing data for the SOILWAT predictions.

We averaged daily soil moisture predictions from SOILWAT from the 5, 25 and 30 and 40 cm soil layers, and daily mean temperature, and then aggregated these seasonally.
